---
output: github_document
---
[![Travis build status](https://travis-ci.org/levisc8/RPadrino.svg?branch=master)](https://travis-ci.org/levisc8/RPadrino)
[![Codecov test coverage](https://codecov.io/gh/levisc8/RPadrino/branch/master/graph/badge.svg)](https://codecov.io/gh/levisc8/RPadrino?branch=master)


```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

## RPadrino

`RPadrino` is a package for interacting with the demography data base of the same name. It will likely consist of wrappers to an SQL data base that itself houses a suite of all known, published integral projection models (IPMs). Additionally, we'll build in some basic tools to re-create the IPMs _as they were created in the original publication_. 

## Scope

The goal of this package is basic data management, exploration, and porting Padrino's internal syntax to [`ipmr`'s](https://levisc8.github.io/ipmr/) syntax. Ideally, one could combine models from here with their own models for synthesis work, using the `proto_ipm` as a common data structure to power the analysis.  

### Installation

You can install RPadrino from github with:

```{r gh-installation, eval = FALSE}

if(!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

remotes::install_github("levisc8/RPadrino")
```

### Usage

A small-ish copy of the Padrino Database is included as a package data set. You can access it with the following code:

```{r eval = FALSE}

data("pdb_ex")

```

The next step is to identify the models we are interested in building. At the moment, there is no functionality to do that - you'll just have to explore the _Metadata_ table on your own. In the not so distant future, there will be a set of functions that actually help query specific columns of interest so the process of exploring the database is less painful.

Once, we've identified the model(s) we want, we can build a list of [`proto_ipm`'s](https://levisc8.github.io/ipmr/articles/proto-ipms.html). This is an intermediate step between the database representation and a set of iteration kernels. Once we have this, we can build an actual model. Below, we extract a specific study, Levin et al. 2019, and construct an IPM from the database object.

```{r eval = FALSE}

# Generate an index to subset the database with

subset_index <- pdb_ex$Metadata$ipm_id[pdb_ex$Metadata$corresponding_author == "Levin"]

# This lapply statement extracts all rows from the database that correspond to our desired study/studies

use_db       <- lapply(pdb_ex,
                       function(x, subset_ind) x[x$ipm_id %in% subset_ind, ],
                       subset_ind = subset_index)

# We can construct a single IPM at a time, or make a list of many IPMs

proto_list   <- make_proto_ipm(use_db, 
                               ipm_id = subset_index[1],
                               det_stoch = "det")


ipm_list     <- make_ipm(proto_list)


```


### Further steps

The package is documented [here]() and the database is documented [here](). 

### Contributing

Please note that the RPadrino project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
