---
output: github_document
---
[![Travis build status](https://travis-ci.com/levisc8/RPadrino.svg?branch=master)](https://travis-ci.com/levisc8/RPadrino)
[![Codecov test coverage](https://codecov.io/gh/levisc8/RPadrino/branch/master/graph/badge.svg)](https://codecov.io/gh/levisc8/RPadrino?branch=master)


```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

## RPadrino

`RPadrino` is a package for interacting with the demography data base of the same name. It will likely consist of wrappers to an SQL data base that itself houses a suite of all known, published integral projection models (IPMs). Additionally, we'll build in some basic tools to re-create the IPMs _as they were created in the original publication_. 

## Scope

The goal of this package is basic data management, exploration, and porting Padrino's internal syntax to [`ipmr`'s](https://levisc8.github.io/ipmr/) syntax. Ideally, one could combine models from here with their own models for synthesis work, using the `proto_ipm` as a common data structure to power the analysis.  

### Installation

You can install RPadrino from github with:

```{r gh-installation, eval = FALSE}

if(!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

remotes::install_github("levisc8/RPadrino")
```

### Contributing

Please note that the RPadrino project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
