---
output: github_document
---

[![R-CMD-check](https://github.com/levisc8/RPadrino/workflows/R-CMD-check/badge.svg)](https://github.com/levisc8/RPadrino/actions) [![Codecov test coverage](https://codecov.io/gh/levisc8/RPadrino/branch/main/graph/badge.svg)](https://codecov.io/gh/levisc8/RPadrino?branch=main)

```{r, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)

library(RPadrino)

data(pdb)

sum_tab <- c(length(unique(pdb$Metadata$species_accepted)), 
             length(unique(pdb$Metadata$apa_citation)),
             nrow(pdb$Metadata))
```

## RPadrino

`RPadrino` is a package for interacting with the Padrino Integral Projection Model database. This database really only exists as an internal dataset in this package right now. The longer term plan is to migrate it to a set of static, version-controlled tables that are hosted on Zenodo (or similar). For now, you can access it by installing the package, and running `data(pdb)`.

The data included in this package are now quality checked and ready for use. A genuine, quality checked release will be tagged when it is actually ready to go. There are `r sum_tab[3]` unique IPMs for `r sum_tab[1]` unique species from `r sum_tab[2]` publications in the internally stored data set.

## Scope

The goal of this package is basic data management, exploration, and porting Padrino's internal syntax to [`ipmr`'s](https://levisc8.github.io/ipmr/) syntax. One can combine models from here with their own models for synthesis work, using the `proto_ipm` as a common data structure to power the analysis.

### Installation

You can install RPadrino from github with:

```{r gh-installation, eval = FALSE}

if(!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

remotes::install_github("levisc8/RPadrino")
```

### Usage

A small-ish copy of the Padrino Database is included as a package data set. You can access it with the following code:

```{r eval = FALSE}

library(RPadrino)

data("pdb")

```

The next step is to identify the models we are interested in building. At the moment, there is no functionality to do that - you'll just have to explore the *Metadata* table on your own. There are accessor functions that allow you to extract metadata columns from the `pdb` object, which may be helpful. See `?pdb_species_accepted` for a list of those.

Once, we've identified the model(s) we want, we can build a list of [`proto_ipm`'s](https://levisc8.github.io/ipmr/articles/proto-ipms.html). This is an intermediate step between the database representation and a set of kernels. Once we have the list of `proto_ipm`'s, we can build an actual model. Below, we extract a specific study, and construct an IPM from the database object.

```{r eval = FALSE}

# We can construct a single IPM at a time, or make a list of many IPMs

proto_list   <- pdb_make_proto_ipm(pdb, 
                                   ipm_id = c("aaaa55", "aaa341", "aaa342"),
                                   det_stoch = "det")

```

Once the list of `proto_ipm`'s is generated, you can either append your own IPMs to it, or you can go ahead to the next chunk. Note that `pdb_make_ipm` will almost always work, but there may still be bugs lurking!

```{r eval = FALSE}

ipm_list <- pdb_make_ipm(proto_list)

lams     <- lambda(ipm_list)

```

### Finding help

The package is documented [here](https://levisc8.github.io/RPadrino/).

### Contributing

Please note that the RPadrino project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
